\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-10 doc/publications/plos16/plos16.tex
changed figures
changes to plos16.tex

\newcommand{\cursedforest}{\textsc{CursedForest}}
\newcommand{\mtry}{{\texttt mtry}}
\newcommand{\ntree}{{\texttt ntree}}

@@

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking joint interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

@@

However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest\ can build a large number of
trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

@@

The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
$H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state
$T$ to a state that takes some information $a$ as given:
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
Variable importance is the sum of IG over all notes and all trees for that variable.

@@

\subsubsection{Synthetic data} 
Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

@@

Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree
whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error-prod.png}
    \caption{OOB classification error} 
    \label{figure:out-of-bag-prediction-error-prod.png} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo-prod.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{figure:rbo-prod.png} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
    the better result, that is, a lower OOB and a higher RBO.}
  \label{figure:synth1}
\end{adjustwidth}
\end{figure}

\marginpar{Does \cursedforest use Gini impurity or information gain? I
  thought it used infomation gain but Piotr mentions Gini}

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in to evaluate ranger \cite{Wright.and.Ziegle.2016},
allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests
by using a range of \mtry\ values.

@@

Firstly we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the
number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\
parameter (presented here as a fraction of the number of variables) [Fig 1 a).

\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
lower and higher values the performance degrades slightly (Fig 1 b). In the latter case the likely cause is
communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

@@

\begin{figure}[tbhp]
\begin{adjustwidth}{-1.00in}{0in}
\begin{tabular}{ll}
a)& b)\\
\includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} & 
\includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png} \\
\end{tabular}
\caption{{\bf Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.} 
  {\bf a)} Time in seconds to build 100 trees for different \mtry\ fractions. 
  {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
\label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

\begin{table}[!ht]
\begin{adjustwidth}{-1.00in}{0in}
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees on 128 CPU cores with varying numbers of variables and samples.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
\cline{3-7}
\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
\hline                                                            
\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{synthetictimingtable}
\end{adjustwidth}
\end{table}

\begin{figure}[tbhp]
\begin{adjustwidth}{-1.00in}{0in}
\begin{tabular}{ll}
a)& b)\\
\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
\end{tabular}
\caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
  {\bf a)} Number of trees build per hour when growing number of variables. 
  {\bf b)} Number of trees build per hour when growing number of samples.}
\label{figure:synth}
\end{adjustwidth}
\end{figure}

>>>>

\newcommand{\cursedforest}{\textsc{CursedForest}\ }
\newcommand{\mtry}{{\texttt mtry\ }}
\newcommand{\ntree}{{\texttt ntree\ }}

@@

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

A tree based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

@@

The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest\ can build a large number of
trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
\marginpar{check this. It also allows us to set mtry to a large number, important when p is large}

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
$B,B^\prime$.
\marginpar{why is this true}  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

@@

The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
\marginpar{this is wrong -- it used Gini}
fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
$H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state
$T$ to a state that takes some information $a$ as given:
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
Variable importance is the sum of IG over all notes and all trees for that variable.

@@

\subsubsection{Synthetic data} 
\marginpar{this is wrong now}
Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

@@

Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree\ 
whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.
\marginpar{this is not true anymore}

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{ OOB (out of bag) classification error and Rank-biased overlap estimates for random
        forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The  rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Each measure is shown
      as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
      the better result, that is, a lower OOB and a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry\ values.

@@

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
number of cores used, \cursedforest displays approximately linear dependency between the execution time and the \mtry
parameter (presented here as a fraction of the number of variables, Fig. \ref{figure:synthetictiming.a}

\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synthetictiming}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
      \label{figure:synthetictiming.a} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
      \caption{Number of trees build per hour when using a growing number of CPU cores.}
    \label{figure:synthetictiming.b}
    \vspace{4ex}
  \end{subfigure} 
\end{adjustwidth}
\end{figure}

@@

\begin{table}[!ht]
\begin{adjustwidth}{-1.00in}{0in}
\centering
\caption{\textbf{
 Time in seconds to build a random forest of 100
  trees on 128 CPU cores with varying numbers of variables and samples.}}
\label{synthetictimingtable}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
\cline{3-7}
\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
\hline                                                            
\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.}}
    \label{figure:synth2}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} 
      \caption{Number of trees build per hour when growing number of variables.} 
      \label{figure:synth2.a} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png}
      \caption{Number of trees build per hour when growing number of samples.}
    \label{figure:synth2.b}
    \vspace{4ex}
  \end{subfigure} 
\end{adjustwidth}
\end{figure}

% \begin{figure}[tbhp]
% \begin{adjustwidth}{-1.00in}{0in}
% \begin{tabular}{ll}
% a)& b)\\
% \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
% \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
% \end{tabular}
% \caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
%   {\bf a)} Number of trees build per hour when growing number of variables. 
%   {\bf b)} Number of trees build per hour when growing number of samples.}
% \label{figure:synth}
% \end{adjustwidth}
% \end{figure}
